Aquifer depletion exacerbates agricultural drought losses in the US High Plains

### 2023-11-01 NatureWaterSubmission/nw_manuscript.tex
separate rnw file created for NW submission. Figures have been modified.
1. font: Helvetica
2. title and legend separately

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage[labelfont=bf]{caption} % \caption*{} will create legend (explanatory text)
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{lineno}
\linenumbers
\usepackage{float}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{authblk}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{plainnat}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Aquifer depletion exacerbates agricultural drought losses in the US High Plains},
  pdfauthor={Taro Mieno; Timothy Foster; Shunkei Kakimoto; Nicholas Brozovic},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

<<init, include = FALSE>>=
knitr::opts_chunk$set(
  echo = FALSE
)
library(tidyverse)
library(data.table)
@

@@

\date{}

\begin{document}
\maketitle
\begin{abstract}
Aquifer depletion poses a major threat to the ability of farmers, food supply chains, and rural economies globally to use groundwater as a means of adapting to climate variability and change. Empirical research has demonstrated the large differences in drought risk exposure that exist between rainfed and irrigated croplands, but previous work commonly assumes water supply for the latter is unconstrained. In this paper, we evaluate how aquifer depletion affects the resilience of irrigated crop production to drought risk using over 30 years of data on historical corn and soybean yields, production areas, and aquifer conditions for the High Plains region in the United States. We show that aquifer depletion reduces the ability of farmers to sustain irrigated crop yields and production areas in years and locations with large growing season water deficits. Our findings demonstrate that drought-related production losses on irrigated croplands increase non-linearly with aquifer depletion, highlighting the need for proactive aquifer conservation interventions to support adaptation and resilience to future increases in rainfall variability under climate change.
\end{abstract}

@@

<<load_results, echo = FALSE>>=
setwd(here::here("Writing"))
all_results <- readRDS("../Results/all_results.rds")
@

\hypertarget{impact-intensive}{%
\subsection{Sensitivity of per-area crop yields to water deficits}\label{impact-intensive}}

@@

<<echo = FALSE>>=
delta_yield_lowest_highest <- 
  all_results[crop == "corn", yield_dif_test][[1]] %>%
  .[comp_q == 3, ] %>%  
  .[which.min(abs(balance -950)), dif_hat] %>%
  round(digits = 2)
@

Figure \ref{fig:yield-dif} shows the difference in yield of the second and third (highest) aquifer quantiles relative to the first (lowest) quantile with the 95\% confidence interval for corn and soybean. For corn, in years with water deficit greater than about 700 mm, the highest aquifer thickness group has statistically significantly higher yields than the lowest aquifer thickness group. At an extreme water deficit level of 950 mm (average water deficit in 2012, the most extreme drought in our historical record), the difference in yields between the smallest and largest aquifer thickness groups is \Sexpr{delta_yield_lowest_highest} tonne/ha. Unlike corn, irrigated soybean yields are not statistically distinguishable among the three aquifer thickness groups. This is likely because soybean is not as water demanding a crop as corn, meaning that even counties with lower aquifer thickness can meet the majority of crop water demands as well yield is not a significantly binding constraint on irrigation management decisions.

<<num_obs_yield_reg, echo = FALSE>>=
nobs_corn <- 
  all_results[crop == "corn", data][[1]] %>% 
  nrow() %>%
  format(big.mark=",")

nobs_soy <- all_results[crop == "soy", data][[1]] %>% 
  nrow() %>%
  format(big.mark=",")
@

@@

<<ir_share, echo = FALSE>>=
corn_share <- all_results[crop == "corn", share_pred_data][[1]]

share_high <-
  corn_share[which.min(abs(sat - 130)), ir_share_hat] %>%
  round(digits = 2)

share_low <-
  corn_share[which.min(abs(sat - 10)), ir_share_hat] %>%
  round(digits = 2) 
@

Figure \ref{fig:ir-share} shows the estimated relationship between aquifer thickness and the share of production area that is irrigated for corn and soybean after accounting for spatio-temporal variations in weather conditions and drought risk exposure across the aquifer. Figure \ref{fig:ir-share} demonstrates that aquifer thickness has meaningfully large impacts on the share of irrigated production for both corn and soybean. As aquifer thickness declines, the share of irrigated acres for corn decreases from around \Sexpr{share_high} (\Sexpr{share_high * 100}\%) for an average county-level aquifer thickness of 100 m to \Sexpr{share_low} (\Sexpr{share_low * 100}\%) for an average county-level aquifer thickness of 10 m (See Appendix \ref{test-dif-share} for formal statistical testings). Reductions in the share of production area under irrigation is also evident for soybean.

<<num_obs_share_reg, echo = FALSE>>=
nobs_corn_share <- 
  all_results[crop == "corn", share_reg_data][[1]] %>% 
  nrow() %>%
  format(big.mark=",")

nobs_soy_share <- 
  all_results[crop == "soy", share_reg_data][[1]] %>% 
  nrow() %>%
  format(big.mark=",")
@

@@

<<res_at_950, echo = FALSE>>=
res_at_950 <-
  all_results[, .(crop, avg_yield_pred_data)] %>%
  unnest(cols = c(avg_yield_pred_data)) %>%
  data.table() %>%
  .[crop == "corn", ] %>%
  .[(abs(balance - 950)) < 5, ]
@

@@

Figure \ref{fig:tot-impact} and Figure \ref{fig:dif-tot-impact} also demonstrate that impact of aquifer thickness on average per-area production is highly non-linear at high seasonal water deficits. At a water deficit of 950 mm, for example, reducing from 100 m aquifer thickness to 70 m (decline of 30 m) would result in only \Sexpr{round(res_at_950[sat == 100 & crop == "corn", avg_yield] - res_at_950[sat == 70 & crop == "corn", avg_yield], digits = 2)} tonne/ha of production loss for corn. This negligible difference in average yield from the decline is due to the insensitivity of the share of irrigated acres to a decline in aquifer thickness when the initial aquifer thickness level is high as can be seen in Figure \ref{fig:ir-share}. Reducing from 70 m aquifer thickness to 40 m (decline of 30 m) would result in \Sexpr{round(res_at_950[sat == 70 & crop == "corn", avg_yield] - res_at_950[sat == 40 & crop == "corn", avg_yield], digits = 2)} tonne/ha of production loss for corn, with a further drop from 40 m aquifer thickness to 10 m (decline of 30 m) leading to \Sexpr{round(res_at_950[sat == 40 & crop == "corn", avg_yield] - res_at_950[sat == 10 & crop == "corn", avg_yield], digits = 2)} tonne/ha of production loss for corn. Similar but less pronounced effects are also observed for soybean. This non-linear impact of aquifer thickness is due to the influence of aquifer thickness on the share of irrigated corn acres (Figure \ref{fig:ir-share}), which implicitly reflects how aquifer thickness alters well yields and, in turn, farmers' ability to effectively and reliably buffer crops against water deficits during the growing season.

@@

These data were combined to form four sets of data for regression analysis: (1) per-area corn yield ($\Sexpr{nobs_corn}$ observations), (2) per-area soybean yield ($\Sexpr{nobs_soy}$ observations), (3) share of irrigated corn ($\Sexpr{nobs_corn_share}$ observations), and (4) share of irrigated soybean ($\Sexpr{nobs_soy_share}$ observations).

@@

\caption*{The solid lines represent per-area irrigated yield responses to water deficits, dissagregated by aquifer thickness level, considering three quantile groups that capture the range of aquifer thickness values in our sample: (i) 9 m to 30.1 m; (ii) 30.1 m to 67.2 m; and (iii) 67.2 m to 215.8 m. The histograms on the top of each panel show the distribution of county-level annual water deficit values included in regressions for each crop. Total number of observations are $\Sexpr{nobs_corn}$ and $\Sexpr{nobs_soy}$ for corn and soybean regressions, respectively.}
\label{fig:yield-response}
\end{figure}

@@

\caption*{The solid red lines represent the difference in estimated per-area irrigated corn (top panels) and soybean (bottom panels) yields for the second and third (largest) aquifer thickness quantiles relative to the first (smallest) aquifer thickness quantile. Shaded area represents 95\% confidence interval.}
\label{fig:yield-dif}
\end{figure}

@@

\begin{figure}[H]
\caption{The estimated impact of aquifer thickness on the share of agricultural production under irrigation for corn and soybean producing counties in the US High Plains}

@@

\label{fig:ir-share}
\caption*{The solid black line represents the estimated share of irrigated acres. Shaded area represents 95\% confidence interval. The histograms on the top of each panel show the distribution of county-level aquifer thickness values included for each crop. Total number of observations are $\Sexpr{nobs_corn_share}$ and $\Sexpr{nobs_soy_share}$ for corn and soybean regressions, respectively.}
\end{figure}

%------------------------
%- Figure 4
%------------------------

@@

\caption{Average productivity of corn and soybean in the US High Plains}

% {\centering \includegraphics[width=6in,height=500px,]{../Results/Figures/g_total_impact}}
\caption*{The solid lines represent average yield at different levels of water deficit and aquifer thickness, considering both per-area yield responses and farmers' choices about irrigated area shares in response to different water deficit and aquifer conditions. See Appendix \ref{reg-conf} for estimated average yield response curves with 95\% confidence intervals displayed, which are omitted here for visual clarity. The histograms on the top of each panel show the distribution of county-level annual water deficit values included in regressions for each crop.}
\label{fig:tot-impact}
\end{figure}

%------------------------
%- Figure 5
%------------------------

@@

\caption{Difference in average yield of corn and soybean for aquifer thickness levels}
% {\centering \includegraphics[width=6in,height=500px,]{../Results/Figures/g_avg_yield_dif}}

\caption*{The solid red lines represent difference in average yield of corn and soybean for aquifer thickness levels of 40 m, 70 m, and 100 m, relative to a baseline aquifer thickness of 10 m. Shaded areas denote 95\% confidence interval, which when greater than zero indicate average yields for the given aquifer thickness level are statistically significantly greater than for an aquifer thickness of 10 m at the 5\% level of significance.}
\label{fig:dif-tot-impact}
\end{figure}

%------------------------
%- Figure 6
%------------------------

@@

\caption{Historical rainfed yield, irrigated yield, and water deficit}

@@

\caption*{The blue solid, red solid, and black dotted lines represent irrigated yield, rainfed yield, and water deficit, respectively.}\label{fig:deficit-yield-hist}
\end{figure}

%------------------------
%- Figure 7
%------------------------

@@

\caption{County-average aquifer thickness in 2016}

@@

\caption{The map shows county-average aquifer thickness in 2016 for each county included in our corn and soybean regression analyses. Counties highlighted in grey are rainfed only, and the red polygon denotes the boundary of the High Plains Aquifer system}\label{fig:sat-map}
\end{figure}

@@

<<sat_ref, echo = FALSE>>=
sat_ref <-
  readRDS(here::here("Shared/Results/all_results.rds")) %>%
  .[1, sat_seq] %>%
  .[[1]] %>%
  .[length(.)]
@

Figure \ref{fig:test-sat-share} presents the difference in the estimated share of irrigated acres at different levels of aquifer thickness relative to the aquifer thickness level of \Sexpr{sat_ref} m, which represents a high level of aquifer thickness for our study area that exerts no constraints on water supply for irrigation. Grey shaded areas represented the 95\% confidence intervals for the estimated relationships. Where this range exceeds the red zero line, this indicates that the difference in aquifer thickness (relative to \Sexpr{sat_ref} m baseline) leads to a statistically significant (at 5\% level) reduction in irrigated production area share.

@@

\caption{Decline in the share of acres that are irrigated as a function of different levels of aquifer thickness relative to a baseline aquifer thickness level of \Sexpr{sat_ref} m (black line). The grey shaded areas represent the 95\% confidence interval for this estimated relationship.}\label{fig:test-sat-share}
\end{figure}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}\usepackage[]{graphicx}\usepackage[]{xcolor}
% maxwidth is the original width if it is less than linewidth
% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
\let\hlipl\hlkwb

\usepackage{framed}
\makeatletter
\newenvironment{kframe}{%
 \def\at@end@of@kframe{}%
 \ifinner\ifhmode%
  \def\at@end@of@kframe{\end{minipage}}%
  \begin{minipage}{\columnwidth}%
 \fi\fi%
 \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
 \colorbox{shadecolor}{##1}\hskip-\fboxsep
     % There is no \\@totalrightmargin, so:
     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
 \MakeFramed {\advance\hsize-\width
   \@totalleftmargin\z@ \linewidth\hsize
   \@setminipage}}%
 {\par\unskip\endMakeFramed%
 \at@end@of@kframe}
\makeatother

\definecolor{shadecolor}{rgb}{.97, .97, .97}
\definecolor{messagecolor}{rgb}{0, 0, 0}
\definecolor{warningcolor}{rgb}{1, 0, 1}
\definecolor{errorcolor}{rgb}{1, 0, 0}
\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

\usepackage{alltt}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage[labelfont=bf]{caption} % \caption*{} will create legend (explanatory text)
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{lineno}
\linenumbers
\usepackage{float}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{authblk}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{plainnat}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Aquifer depletion exacerbates agricultural drought losses in the US High Plains},
  pdfauthor={Taro Mieno; Timothy Foster; Shunkei Kakimoto; Nicholas Brozovic},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

\date{}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}
\maketitle
\begin{abstract}
Aquifer depletion poses a major threat to the ability of farmers, food supply chains, and rural economies globally to use groundwater as a means of adapting to climate variability and change. Empirical research has demonstrated the large differences in drought risk exposure that exist between rainfed and irrigated croplands, but previous work commonly assumes water supply for the latter is unconstrained. In this paper, we evaluate how aquifer depletion affects the resilience of irrigated crop production to drought risk using over 30 years of data on historical corn and soybean yields, production areas, and aquifer conditions for the High Plains region in the United States. We show that aquifer depletion reduces the ability of farmers to sustain irrigated crop yields and production areas in years and locations with large growing season water deficits. Our findings demonstrate that drought-related production losses on irrigated croplands increase non-linearly with aquifer depletion, highlighting the need for proactive aquifer conservation interventions to support adaptation and resilience to future increases in rainfall variability under climate change.
\end{abstract}

@@

\hypertarget{impact-intensive}{%
\subsection{Sensitivity of per-acre crop yields to water deficits}\label{impact-intensive}}

@@

Figure \ref{fig:yield-dif} shows the difference in yield of the second and third (highest) aquifer quantiles relative to the first (lowest) quantile with the 95\% confidence interval for corn and soybean. For corn, in years with water deficit greater than about 700 mm, the highest aquifer thickness group has statistically significantly higher yields than the lowest aquifer thickness group. At an extreme water deficit level of 950 mm (average water deficit in 2012, the most extreme drought in our historical record), the difference in yields between the smallest and largest aquifer thickness groups is 1.31 tonne/ha. Unlike corn, irrigated soybean yields are not statistically distinguishable among the three aquifer thickness groups. This is likely because soybean is not as water demanding a crop as corn, meaning that even counties with lower aquifer thickness can meet the majority of crop water demands as well yield is not a significantly binding constraint on irrigation management decisions.

@@

Figure \ref{fig:ir-share} shows the estimated relationship between aquifer thickness and the share of production area that is irrigated for corn and soybean after accounting for spatio-temporal variations in weather conditions and drought risk exposure across the aquifer. Figure \ref{fig:ir-share} demonstrates that aquifer thickness has meaningfully large impacts on the share of irrigated production for both corn and soybean. As aquifer thickness declines, the share of irrigated acres for corn decreases from around 0.89 (89\%) for an average county-level aquifer thickness of 100 m to 0.7 (70\%) for an average county-level aquifer thickness of 10 m (See Appendix \ref{test-dif-share} for formal statistical testings). Reductions in the share of production area under irrigation is also evident for soybean.

@@

@@

Figure \ref{fig:tot-impact} and Figure \ref{fig:dif-tot-impact} also demonstrate that impact of aquifer thickness on average per-area production is highly non-linear at high seasonal water deficits. At a water deficit of 950 mm, for example, reducing from 100 m aquifer thickness to 70 m (decline of 30 m) would result in only 0.15 tonne/ha of production loss for corn. This negligible difference in average yield from the decline is due to the insensitivity of the share of irrigated acres to a decline in aquifer thickness when the initial aquifer thickness level is high as can be seen in Figure \ref{fig:ir-share}. Reducing from 70 m aquifer thickness to 40 m (decline of 30 m) would result in 0.95 tonne/ha of production loss for corn, with a further drop from 40 m aquifer thickness to 10 m (decline of 30 m) leading to 1.45 tonne/ha of production loss for corn. Similar but less pronounced effects are also observed for soybean. This non-linear impact of aquifer thickness is due to the influence of aquifer thickness on the share of irrigated corn acres (Figure \ref{fig:ir-share}), which implicitly reflects how aquifer thickness alters well yields and, in turn, farmers' ability to effectively and reliably buffer crops against water deficits during the growing season.

@@

These data were combined to form four sets of data for regression analysis: (1) per-acre corn yield ($8,814$ observations), (2) per-acre soybean yield ($5,997$ observations), (3) share of irrigated corn ($2,464$ observations), and (4) share of irrigated soybean ($1,553$ observations).

@@

\caption*{Irrigated yield responses to water deficits are dissagregated by aquifer thickness level, considering three quantile groups that capture the range of aquifer thickness values in our sample: (i) 9 m to 30.1 m; (ii) 30.1 m to 67.2 m; and (iii) 67.2 m to 215.8 m. The histograms on the top of each panel show the distribution of county-level annual water deficit values included in regressions for each crop. Total number of observations are $8,814$ and $5,997$ for corn and soybean regressions, respectively.}
\label{fig:yield-response}
\end{figure}

@@

\caption*{The solid red lines represent the difference in estimated per-acre irrigated corn (top panels) and soybean (bottom panels) yields for the second and third (largest) aquifer thickness quantiles relative to the first (smallest) aquifer thickness quantile. Shaded area represents 95\% confidence interval.}
\label{fig:yield-dif}
\end{figure}

@@

\begin{figure}[H]
\caption{The estimated impact of aquifer thickness on the share of agricultural production under irrigation for corn and soybean producing counties in the US High Plains.}

@@

\label{fig:ir-share}
\caption*{The solid black line represents the estimated share of irrigated acres. Shaded area represents 95\% confidence interval. The histograms on the top of each panel show the distribution of county-level aquifer thickness values included for each crop. Total number of observations are $2,464$ and $1,553$ for corn and soybean regressions, respectively.}
\end{figure}

@@

% {\centering \includegraphics[width=6in,height=500px,]{../Results/Figures/g_total_impact}}

\caption{Average productivity of corn and soybean in the US High Plains for different levels of water deficit and aquifer thickness, considering both per-area yield responses and farmers' choices about irrigated area shares in response to different water deficit and aquifer conditions. See Appendix \ref{reg-conf} for estimated average yield response curves with 95\% confidence intervals displayed, which are omitted here for visual clarity.}\label{fig:tot-impact}
\end{figure}

@@

% {\centering \includegraphics[width=6in,height=500px,]{../Results/Figures/g_avg_yield_dif}}

\caption{Difference in average yield of corn and soybean for aquifer thickness levels of 40 m, 70 m, and 100 m, relative to a baseline aquifer thickness of 10 m. Shaded areas denote 95\% confidence interval, which when greater than zero indicate average yields for the given aquifer thickness level are statistically significantly greater than for an aquifer thickness of 10 m at the 5\% level of significance.}\label{fig:dif-tot-impact}
\end{figure}

@@

@@

\caption{Historical rainfed yield, irrigated yield, and water deficit. The blue solid, red solid, and black dotted lines represent irrigated yield, rainfed yield, and water deficit, respectively.}\label{fig:deficit-yield-hist}
\end{figure}

@@

@@

\caption{Map showing county-average aquifer thickness in 2016 for each county included in our corn and soybean regression analyses. Counties highlighted in grey are rainfed only, and the red polygon denotes the boundary of the High Plains Aquifer system}\label{fig:sat-map}
\end{figure}

@@

Figure \ref{fig:test-sat-share} presents the difference in the estimated share of irrigated acres at different levels of aquifer thickness relative to the aquifer thickness level of 100 m, which represents a high level of aquifer thickness for our study area that exerts no constraints on water supply for irrigation. Grey shaded areas represented the 95\% confidence intervals for the estimated relationships. Where this range exceeds the red zero line, this indicates that the difference in aquifer thickness (relative to 100 m baseline) leads to a statistically significant (at 5\% level) reduction in irrigated production area share.

@@

\caption{Decline in the share of acres that are irrigated as a function of different levels of aquifer thickness relative to a baseline aquifer thickness level of 100 m (black line). The grey shaded areas represent the 95\% confidence interval for this estimated relationship.}\label{fig:test-sat-share}
\end{figure}
